Supplementary Material

### 2025-02-12 response_to_reviewers.qmd

---
title: ""
format: 
  docx:
    reference-doc: "ref.docx"
bibliography: "misc/ref.bib"
csl: "misc/scientific-reports.csl"
editor_options: 
  chunk_output_type: console
execute: 
  echo: false
  warning: false
  error: false
---

# Response to Reviewers

## Overview

We thank the Reviewers for their valuable feedback and constructive comments up to this point, which have helped us to significantly improve the quality and clarity of our manuscript. Below, we provide a detailed response to the comments from reviewer 2.

## Reviewer 2

**Comment**: "A lot of determinism has been previously demonstrated, where the respiration is spontaneous, e.g. as in Physica A, 283, no. 3-4 (2000): 451-461; and Philos. Trans. R. Soc. A, 371, no. 1997 (2013): 20110622;"

**Response**: We appreciate the reviewer’s suggestion and acknowledge the importance of prior deterministic models in characterizing cardio-respiratory interactions. In response, we have included a discussion in our manuscript that situates our model within this deterministic framework, highlighting how logistic functions effectively capture structured, time-dependent fluctuations in RR intervals. We have also cited the suggested references to acknowledge their contributions to the field.

**Comment**: "Similarly, if a complementary approach is taken, then the fluctuations of the cardio-respiratory interactions have been studied also through a stochastic approach, eg. as in Eur. Phys. J. B, 65 (2008): 425-433"

**Response**: We recognize the significance of stochastic modeling in capturing the inherent variability of cardio-respiratory interactions. While our current model is primarily deterministic, it does account for inter-individual variability, which aligns with aspects of stochastic approaches. We have now included a reference to prior stochastic studies and briefly discussed how future extensions of our model could incorporate stochastic elements to enhance its adaptability in more variable physiological conditions.

**Comment**: "The authors could reflect also these important works, if they have further opportunity in the revision process."

**Response**: We appreciate the reviewer’s insightful suggestion. In response, we have revised the discussion section to include references to both deterministic and stochastic modeling approaches, clarifying how our work aligns with and builds upon these perspectives. We believe this addition strengthens the manuscript by providing a broader context for our modeling approach.

### 2025-02-12 paper.qmd
2nd round of reviews
- reviewer 2 ask to add a bit more dicussion on stochastic and deterministic models.
- editor and reviewer 1 had no more comments.
- hope this is the final round of reviews!

**Title**: Enhancing Cardiovascular Monitoring: A Non-Linear Approach to RR Interval Dynamics in Exercise and Recovery.

**Authors**: Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.

::: {custom-style="Afiliation"}
^1^ Centro Asistencial Docente e Investigación (CADI-UMAG), University of Magallanes, Punta Arenas, Chile.

^2^ Kinesiology Department, University of Magallanes, Punta Arenas, Chile.

^3^ School of Medicine, University of Magallanes, Punta Arenas, Chile.

@@

Cristian Núñez-Espinosa, School of Medicine, University of Magallanes, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411.

@@

This work aimed to develop and validate a novel non-linear model to characterize RR interval (RRi) dynamics throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches. Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi dynamics, with parameter estimation achieved via Hamiltonian Monte Carlo. Sobol sensitivity analysis identified baseline RRi ($\alpha$) and recovery proportion ($c$) as the primary drivers of variability, underscoring their critical roles in autonomic regulation and resilience. Validation against real-world RRi data demonstrated robust model performance, accurately reflecting autonomic recovery and exercise-induced fluctuations. By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring in athletic contexts to optimize performance and recovery. These findings highlight the model's ability to provide precise, physiologically relevant assessments of autonomic function, paving the way for its use in personalized health monitoring and performance optimization across diverse populations.

@@

Current research has extensively examined the mechanisms underlying cardiac autonomic dynamics in response to exercise and their links to health-related quality of life and cardiovascular disease risk [@eser2022exercise;@castillo2023validity;@mabe2024associations].

In this context, the study of R-R intervals (RRi), defined as the time intervals between heartbeats and their link with exercise, has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices]. Hence, analyzing the temporal dynamics of RRi in response to exercise can provide valuable insights into how the cardiovascular system adapts to physical stressors, such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].

Unlike heart rate variability (HRV), which aggregates autonomic responses over time, RRi analysis provides a more granular, direct view of cardiac electrical activity during or immediately following exercise, particularly in older adults [@mongin2022hrv; @castillo2023validity; @mabe2024associations].

Thus, understanding RRi fluctuations in response to exercise is particularly relevant during dynamic exercise periods, where the autonomic nervous system (ANS) shifts between parasympathetic withdrawal and sympathetic activation [@boettger2010heart]. Therefore, directly modeling RRi dynamics, rather than relying on broader HRV metrics, allows for a direct assessment of physiological markers of autonomic adaptation to stress [@hautala2003short]. This approach is valuable for identifying recovery patterns and understanding cardiovascular reactivity across individuals with various fitness levels [@mongin2022hrv].

Modeling the RRi behavior has traditionally been approached by leveraging linear regression and time-series analysis [@lian2011simple]. However, this oppose significant challenges, like oversimplifying overall exercise dynamics without capturing the intricacies of exercise-induced intricate transitions in RRi, especially under intense exertion and recovery phases [@gronwald2019effects]. More recently, advanced non-linear approaches have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated;@lee2023real;@berrahou2024arrhythmia]. However, many more advanced alternatives fail to generate physiologically meaningful model parameters without directly linking to biological processes [@gronwald2019non].

Alongside this line of inquiry, recent studies have begun exploring non-linear models for RRi dynamics, recognizing their potential to capture the complexity of cardiovascular response to exercise [@da2024post]. Exponential decay models, for example, have been proposed to describe RRi recovery [@sahoo2024unanticipated], while logistic functions have been used to model the gradual return to baseline after high-intensity exercise [@molkkari2020online; @gronwald2019non]. These models offer advantages over traditional HRV metrics by providing a more detailed understanding of the cardiovascular system's response to exercise [@wu2003nonlinear]. However, despite these advancements, few models are specifically designed to capture real-time RRi fluctuations, and even fewer consider physiologically meaningful parameters that allow modeling individual variability across a myriad of exercise dimensions [@kanniainen2023estimation].

Considering the current state of the art, there is a clear need to develop a non-linear model that accurately represents RRi's non-linear transitions during exercise and recovery, where the estimation of clinically significant parameters is compelling. Such a model would need to offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy]. This model must be complex enough to capture the non-linear, exercise-induced cardiovascular dynamics. But, simultaneously, it has to be simple enough to provide practical and significant model parameters related to observed physiological processes.

Hence, the primary objective of this paper is to present a novel non-linear model that characterizes continuous RRi transitions from rest to exercise and recovery. This model is designed to capture the non-linear changes in RRi, providing meaningful parameters that can enhance our understanding of the physiological processes underlying cardiovascular adaptation to exercise.

@@

To further assess the performance of the proposed model, real-world RRi data were analyzed in addition to the synthetic data generated through simulation. This dataset was derived from a cohort participating in the FONDECYT Project No. 11220116, funded by the Chilean National Association of Research and Development (ANID). Ethical approval was granted by the Ethics Committee of the University of Chile (ACTA No. 029-18/05/2022) and the Ethics Committee of the University of Magallanes (No. 008/SH/2022).

The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity]. Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, which is capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.

@@

This real-world dataset provided a critical context for validating the model's predictive capability against observed physiological responses, offering a robust foundation for understanding RRi dynamics under physical activity conditions.

@@

The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package (v2.21.0), which employs the Stan probabilistic programming language. Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes, targeted at a minimum of 1,000 for each parameter [@brms]. Trace plots were inspected to confirm stable mixing. These diagnostics collectively confirmed reliable posterior estimates for each parameter.

@@

To enhance the exploration of parameter space, we performed a two-stage analysis: We assessed parameter values at the individual level, which we then used to estimate  population-level parameters.

@@

$$
y_{i,t} = \frac{\text{RRi}_{i,t} - \bar{\text{RRi}_i}}{S_{\text{RRi}_i}}
$${#eq-ind-std}
This standardization allowed the model to focus on relative changes in RRi dynamics, independent of individual baseline differences.

@@

$$
y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-c_i \cdot \beta_i}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
$${#eq-indiv-level}
where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.

@@

This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates of subject- and population-level model parameters.

@@

The primary performance metrics, estimated for each subject, included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject. Bootstrap resampling across each metric was performed to estimate the mean performance of the model and corresponding quantile-based 95% CI.

Also, residual analysis were conducted to evaluate the model's accuracy in capturing RRi dynamics. Residuals were defined as the difference between observed and predicted RRi values. These residuals were analyzed for temporal structure and partial autocorrelation to ensure that no systematic patterns remained in the errors. This indicates that the model has sufficiently captured the underlying dynamics of the RRi response to exercise.

@@

We implemented a Sobol sensitivity analysis using Monte Carlo simulations to assess the sensitivity of model parameters influencing RRi over time. Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise [@zhang2015sobol].

To compute $S_{\text{ind}}$, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs). For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes. The 95% CI parameter values estimated from HMC-NUTS were then used as input ranges for $S_{\text{ind}}$ computation. Finally, the mean values of $S_{\text{ind}}$ over the 20-minute time span for each model parameter were estimated and reported, with their corresponding 95% CI using normal approximation based on estimated standard errors (SE).

@@

RRi dynamics in response to exercise tends to follow a U-shaped form. The initial decrease in RRi is associated with exercise onset, related to an increased heart rate. After exercise cessation, an opposite increase in RRi is observed, which is associated with the cardiovascular recovery phase. In both cases, the drop and recovery phases occurs at different rates; some individuals experience a quick recovery in RRi after exercise; however, in some others, this slope is less steep. Additionally, the new baseline reached following exercise cessation is often below the RRi baseline before exercise.

These hallmarks of RRi dynamics in response to exercise, highlight the complex and non-linear behavior of the cardiovascular response in the context of both rest and exercise conditions. [Figure 1](#fig-rri-example) shows an example of RRi record data.

<!---- FIGURE 1 ---->

@@

The process of deriving the final equation for modeling RRi fluctuations was guided by an iterative exploration of mathematical functions capable of capturing the observed dynamics. Initially, exponential and logarithmic functions were considered due to their simplicity and wide applicability in describing temporal changes. Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset. In contrast, logarithmic functions were explored for their capacity to describe asymptotic behaviors observed in some physiological variables.

@@

We shifted to logistic functions to address these limitations, which inherently model sigmoidal transitions. Logistic functions introduce parameters for growth rate and inflection point, allowing for precise control over the shape and timing of the transition between dynamic states. By using two coupled logistic functions, one to represent the initial decrease in RRi and a second inverted logistic function describing the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.

This approach provided a biologically plausible representation, with parameters that directly correspond to identifiable physiological features, such as the rate of adaptation and recovery, the time to peak response, and the extent of deviation from baseline. The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured the qualitative and quantitative aspects of RRi dynamics.

@@

This model includes two logistic functions representing the RRi dynamics across exercise and recovery phases. The first logistic term models the decrease in RRi during exercise, where the parameter $\beta$ denotes the magnitude of this decline. The rate of decrease is governed by $\lambda$, while $\tau$ represents the onset of the RRi decrease or the time the physiological shift begins.

The second logistic term accounts for RRi recovery post-exercise. Here, $c$ scales the magnitude of recovery relative to the initial decline represented by $\beta$, capturing the proportion of the decline regained during recovery. The rate at which RRi returns to baseline is controlled by $\phi$, and $\delta$ indicates the lag following the cessation of exercise, marking the beginning of recovery, respectively.

Additionally, the dynamics of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions denoted as $f_1(t)$ and $f_2(t)$. The function $f_1(t)$ models the initial decay in RRi following the initiation of exercise while $f_2(t)$ characterizes the recovery phase after exercise cessation.

@@

<!---- FIGURE 2 ---->

@@

The sample used to assess RRi dynamics consists of 272 subjects selected from a local community of elderly individuals. The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)

<!---- TABLE 1 ---->

An initial graphical exploration of RRi dynamics (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated with exercise-induced cardiovascular stress. However, greater variability across individuals in post-exercise recovery can be observed.

<!---- FIGURE 3 ---->

@@

Simulated standardized RRi dynamics based on prior parameter distributions are shown in [Figure 4](#fig-prior-sim).

<!---- FIGURE 4 ---->

@@

<!---- TABLE 2 ---->

@@

# Figures

>>>>

**Title**: Enhancing Cardiovascular Monitoring: A Non-Linear Model for Characterizing RR Interval Fluctuations in Exercise and Recovery.

**Authors**: Matías Castillo-Aguilar^1^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.

::: {custom-style="Afiliation"}
^1^ Centro Asistencial Docente e Investigación (CADI-UMAG), Universidad de Magallanes, Punta Arenas, Chile.

^2^ Departamento de Kinesiología, Universidad de Magallanes, Punta Arenas, Chile.

^3^ Escuela de Medicina, Universidad de Magallanes, Punta Arenas, Chile.

@@

Cristian Núñez-Espinosa, Escuela de Medicina, Universidad de Magallanes, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411.

@@

This work aimed to develop and validate a novel non-linear model to characterize RR interval (RRi) time-dependent fluctuations throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches. Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi time-dependent fluctuations, with parameter estimation achieved via Hamiltonian Monte Carlo. Sobol sensitivity analysis identified baseline RRi ($\alpha$) and recovery proportion ($c$) as the primary drivers of variability, underscoring their critical roles in autonomic regulation and resilience. Validation against real-world RRi data demonstrated robust model performance ($R^2$ = 0.868, CI~95%~[0.834, 0.895] and 
Root Mean Square Error [RMSE] = 32.6 ms, CI~95%~[30.01, 35.77]), accurately reflecting autonomic recovery and exercise-induced fluctuations. By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring in athletic contexts to optimize performance and recovery. These findings highlight the model's ability to provide precise, physiologically relevant assessments of autonomic function, paving the way for its use in personalized health monitoring and performance optimization across diverse populations.

@@

The human cardiovascular system exhibits intricate dynamic responses to physical exertion, reflecting the complex interplay between the autonomic nervous system (ANS) and cardiac function. Understanding these time-dependent fluctuations is crucial for assessing physiological adaptation to exercise, optimizing athletic performance, and evaluating cardiovascular health [@eser2022exercise; @castillo2023validity; @mabe2024associations]. R-R intervals (RRi), representing the beat-to-beat time intervals between successive heartbeats, provide a direct, high-resolution reflection of cardiac electrical activity. Unlike aggregated measures of heart rate variability (HRV), which summarize autonomic activity over longer periods and can mask transient fluctuations, RRi analysis offers a granular, beat-to-beat perspective on autonomic modulation during exercise and recovery [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].

This granular perspective is particularly relevant in dynamic exercise scenarios, where rapid shifts in autonomic balance occur, and in specific populations such as older adults, where age-related changes in autonomic function may influence cardiac responses [@mongin2022decrease; @castillo2023validity; @mabe2024associations]. Analyzing RRi allows for examining immediate cardiac responses to exercise-induced stress, providing valuable insights into the efficiency and adaptability of the cardiovascular system.

While many studies have investigated cardiovascular responses to exercise using quasi-stationary protocols, simplifying analysis by minimizing non-stationarities [@storniolo2021symbolic; @porta2018relevance], these approaches may not fully capture physiological responses' dynamic and continuous nature during real-world activities. Although traditional linear methods like time-series analysis and linear regression have been employed to model RRi behavior [@lian2011simple], they often fall short in capturing the complex, non-linear time-dependent fluctuations of RRi transitions, particularly during periods of intense exertion and the subsequent recovery phase [@gronwald2019effects]. This limitation is significant because the ANS undergoes rapid and non-linear shifts between parasympathetic withdrawal and sympathetic activation during exercise, resulting in intricate RRi fluctuations that linear models cannot adequately represent by their nature [@boettger2010heart]. These rapid autonomic adjustments, including vagal tone and sympathetic outflow changes, contribute to the non-linear patterns observed in RRi data [@kristal1995heart; @thayer2010relationship]. Consequently, these simplified models may miss critical physiological information related to cardiovascular adaptation, such as the speed and extent of recovery [@hautala2003short].

Model-based approaches, particularly those employing exponential functions, have been widely used to estimate heart rate and RRi recovery time constants after exercise [@imai1994vagally; @borresen2008autonomic; @pierpont2000heart; @pierpont2004assessing; @buchheit2007parasympathetic; @peccanha2014heart]. While these models provide valuable insights into recovery kinetics, they often focus on specific phases of the exercise-recovery cycle. They may not fully capture the continuous transitions in RRi from rest to exercise and back to baseline. Furthermore, these models often rely on simplifying assumptions about the underlying physiological mechanisms, which may limit their ability to represent individual variability across different exercise intensities and populations accurately. Other models, like advanced non-linear approaches, have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated; @lee2023real; @berrahou2024arrhythmia]. More advanced techniques, such as non-linear mode decomposition [@iatsenko2015nonlinear; @iatsenko2016extraction], dynamical modeling [@kralemann2013vivo; @stankovski2012inference], and the explicit consideration of non-autonomous dynamics [@clemson2014discerning; @lehnertz2024time], have also been applied to analyze physiological time series.

However, many of these existing non-linear models' lack of a direct link to underlying physiological processes is a significant limitation. While they may provide a better fit to the observed data, they often lack clear physiological interpretability, limiting their clinical utility and hindering a deeper understanding of the mechanisms driving RRi changes [@gronwald2019non; @bacopoulou2021entropy; @da2024post]. Furthermore, few models are designed to capture the continuous, beat-to-beat transitions in RRi throughout the entire rest-exercise-recovery cycle while simultaneously providing physiologically meaningful parameters that can explain individual variability across diverse exercise intensities, durations, and populations [@kanniainen2023estimation]. This gap hinders a comprehensive understanding of how individuals adapt to exercise and how these adaptations might differ based on age, fitness level, or underlying health status. For example, understanding how RRi time-dependent fluctuations differ between trained athletes and sedentary individuals during and after exercise could provide valuable insights for personalized training programs and rehabilitation strategies.

Therefore, this paper introduces a novel non-linear model designed to characterize the continuous RRi transitions from rest to exercise and recovery. This model aims to address the limitations of existing approaches by (1) accurately capturing the non-linear time-dependent fluctuations of RRi fluctuations throughout the entire rest-exercise-recovery cycle, providing a more complete picture of cardiovascular responses to exercise, and (2) providing physiologically interpretable parameters that reflect the underlying autonomic mechanisms, allowing for a more mechanistic understanding of individual adaptations. By focusing on these key aspects, this model offers a more detailed and physiologically relevant understanding of cardiovascular adaptation to exercise compared to traditional HRV metrics, with potential applications in personalized exercise prescription, performance monitoring, and clinical assessment of cardiovascular health.

@@

To further assess the proposed model's performance, real-world RRi data were analyzed in addition to the synthetic data generated through simulation. The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity].

### Subjects

Participants were recruited from a local community. Subjects were included if (i) they were aged 60 years or older; (ii) were permanently residing in the Magallanes and Chilean Antarctic region; (iii) had a percentage greater than 60% on the Karnofsky Performance Status Scale, which allowed us to work with older people who had a state of autonomy necessary to carry out the study tests; (iv) absence of the following diagnosis: diabetic neuropathy; use of pacemakers; clinical depression; cognitive or motor disability; and dementia. The exclusion criteria were: (i) consumption of beta-blockers during the study, (ii) taking drugs or stimulant substances within 12 hours before the cardiac assessment, and (iii) having some degree of motor disability that prevented participants from moving around. No participants met the exclusion criteria. This dataset was derived from a cohort participating in the FONDECYT Project No. 11220116, funded by the Chilean National Association of Research and Development (ANID). Ethical approval was granted by the Ethics Committee of the University of Chile (ACTA No. 029-18/05/2022) and the Ethics Committee of the University of Magallanes (No. 008/SH/2022).

### Exercise protocol

The exercise protocol consisted of the continuous measurement of RRi before, during, and immediately after the 2-minute step test, which is a part of the Senior Fitness Test protocol [@rikli2013senior]. This functional cardiorespiratory test required each subject to march in place as many times as possible for 2 minutes. The participants were monitored throughout the assessment using cardiovascular measures (i.e., heart rate and blood pressure) to prevent adverse events during the exercise protocol. The evaluation protocol was estimated to last approximately 20 minutes for each subject. None of the participants expressed discomfort during the evaluation. Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.

### Preprocessing of RRi data

@@

This real-world dataset provided a critical context for validating the model's predictive capability against observed physiological responses, offering a robust foundation for understanding RRi time-dependent fluctuations under physical activity conditions.

@@

The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package (v2.21.0), which employs the Stan probabilistic programming language. Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes targeted at a minimum of 1,000 for each parameter [@brms]. Trace plots were inspected to confirm stable mixing. These diagnostics collectively confirmed reliable posterior estimates for each parameter.

@@

To enhance the exploration of parameter space, we performed a two-stage analysis: We assessed parameter values at the individual level, which we then used to estimate  population-level parameters. This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates of subject- and population-level model parameters.

@@

$$
y_{i,t} = \frac{\text{RRi}_{i,t} - \bar{\text{RRi}_i}}{S_{\text{RRi}_i}}
$${#eq-ind-std}

This standardization allowed the model to focus on relative changes in RRi time-dependent fluctuations independent of individual baseline differences.

@@

$$
y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-c_i \cdot \beta_i}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
$${#eq-indiv-level}

where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.

@@

@@

The primary statistical performance metrics, estimated for each subject, included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject. Bootstrap resampling across each metric was performed to estimate the mean performance of the model and corresponding quantile-based 95% CI.

Also, residual analysis were conducted to evaluate the model's accuracy in capturing RRi time-dependent fluctuations. Residuals were defined as the difference between observed and predicted RRi values. These residuals were analyzed for temporal structure and partial autocorrelation to ensure that no systematic patterns remained in the errors. This indicates that the model has sufficiently captured the underlying time-dependent fluctuations of the RRi response to exercise.

@@

We implemented a Sobol sensitivity analysis using Monte Carlo simulations to assess the sensitivity of model parameters influencing RRi over time. Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi time-dependent fluctuations in response to exercise [@zhang2015sobol].

To compute $S_{\text{ind}}$, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs). For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes. The 95% CI parameter values estimated from HMC-NUTS were then used as input ranges for $S_{\text{ind}}$ computation. Finally, the mean values of $S_{\text{ind}}$ over the 20-minute time span for each model parameter were estimated and reported, with their corresponding 95% CI using a normal approximation based on estimated standard errors (SE).

@@

RRi time-dependent fluctuations in response to exercise tend to follow a U-shaped form. The initial decrease in RRi is associated with exercise onset and an increased heart rate. After exercise cessation, an opposite increase in RRi is observed, associated with the cardiovascular recovery phase. In both cases, the drop and recovery phases occur at different rates; some individuals experience a quick recovery in RRi after exercise; however, in some others, this slope is less steep. Additionally, the new baseline reached following exercise cessation is often below the RRi baseline before exercise.

These hallmarks of RRi time-dependent fluctuations in response to exercise highlight the complex and non-linear behavior of the cardiovascular response in the context of rest and exercise conditions. [Figure 1](#fig-rri-example) shows an example of RRi record data.

<!---- FIGURE 1 ---->
![](figures/figure-1.png)
[**Figure 1**]{#fig-rri-example}. Example data of RRi recordings of 6 subjects over a 20-minute rest-exercise-recovery protocol in a sample of elderly individuals. The subject-level data shows the inter-individual variability of RRi time-dependent fluctuations in response to exercised-induced cardiovascular stress, with similar behavior and recovery trajectories over time.
<!---- FIGURE 1 ---->

@@

The process of deriving the final equation for modeling RRi fluctuations was guided by an iterative exploration of mathematical functions capable of capturing the observed time-dependent fluctuations. Initially, exponential and logarithmic functions were considered due to their simplicity and broad applicability in describing temporal changes. Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset. In contrast, logarithmic functions were explored for their capacity to describe asymptotic behaviors observed in some physiological variables.

@@

We shifted to logistic functions to address these limitations, which inherently model sigmoidal transitions. Logistic functions introduce parameters for growth rate and inflection point, allowing for precise control over the shape and timing of the transition between dynamic states. By using two coupled logistic functions, one representing the initial decrease in RRi and a second inverted logistic function describing the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.

This approach provided a biologically plausible representation, with parameters that directly correspond to identifiable physiological features, such as the rate of adaptation and recovery, the time to peak response, and the extent of deviation from baseline. The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured the qualitative and quantitative aspects of RRi time-dependent fluctuations.

@@

This model includes two logistic functions representing the RRi time-dependent fluctuations across exercise and recovery phases. The first logistic term models the decrease in RRi during exercise, where the parameter $\beta$ denotes the magnitude of this decline. The rate of decrease is governed by $\lambda$, while $\tau$ represents the onset of the RRi decrease or the time the physiological shift begins.

The second logistic term accounts for RRi recovery post-exercise. Here, $c$ scales the magnitude of recovery relative to the initial decline represented by $\beta$, capturing the proportion of the decline regained during recovery. The rate at which RRi returns to baseline is controlled by $\phi$, and $\delta$ indicates the lag following the cessation of exercise, marking the beginning of recovery.

Additionally, the time-dependent fluctuations of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions denoted as $f_1(t)$ and $f_2(t)$. The function $f_1(t)$ models the initial decay in RRi following the initiation of exercise while $f_2(t)$ characterizes the recovery phase after exercise cessation.

@@

<!---- FIGURE 2 ---->
![](figures/figure-2.png)
[**Figure 2**]{#fig-linear-constituents}. The RRi time-dependent fluctuations in response to exercise are expressed as a linear combination of model constituents based on the baseline RRi $\alpha$ and two logistic functions, denoted $f_1(t)$ and $f_2(t)$, respectively. The vertical dashed lines represent the time at which the exercise and recovery onset given by $\tau = 5$ and $\delta = 2$.
<!---- FIGURE 2 ---->

@@

The sample used to assess RRi time-dependent fluctuations consists of 272 subjects selected from a local community of elderly individuals. The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)

<!---- TABLE 1 ---->
|Characteristic |Overall        |Female         |Male           |
|:--------------|--------------:|--------------:|--------------:|
|Sex            |—              |217 (79.8%)    |55 (20.2%)     |
|Age            |71.14 ± 6.03   |70.73 ± 6.27   |72.73 ± 4.7    |
|SBP (mm hg)    |130.23 ± 17.07 |129.58 ± 17.37 |132.8 ± 15.69  |
|DBP (mm hg)    |77.1 ± 9.58    |76.68 ± 9.83   |78.75 ± 8.4    |
|MAP (mm hg)    |94.81 ± 10.69  |94.31 ± 10.95  |96.76 ± 9.45   |
|PP (mm hg)     |53.14 ± 14.07  |52.9 ± 14.26   |54.05 ± 13.38  |
|BMI            |30.66 ± 5.43   |30.7 ± 5.64    |30.53 ± 4.53   |
|Weight (kg)    |75.06 ± 14.23  |73.88 ± 14.09  |79.69 ± 13.95  |
|Height (cm)    |156.56 ± 9.18  |155.29 ± 8.46  |161.55 ± 10.24 |

[**Table 1**]{#tbl-sample-characteristics}. Sample characteristics from which continuous RRi monitoring data was collected during the rest-exercise-rest protocol. Data is presented as Mean ± standard deviation (SD). SBP, systolic blood pressure; DBP, diastolic blood pressure; MAP, mean arterial pressure; PP, pulse pressure; BMI, body mass index.
<!---- TABLE 1 ---->

An initial graphical exploration of RRi time-dependent fluctuations (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes, associated with exercise-induced cardiovascular stress. However, greater variability across individuals in post-exercise recovery can be observed.

<!---- FIGURE 3 ---->
![](figures/figure-3.png)
[**Figure 3**]{#fig-2d-kernel-density}. (A) Mean and SD from each subject's RRi recordings were used for the standardization process. (B) 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol. Darker colors indicate greater probability density. The contrary can be said about lighter colors.
<!---- FIGURE 3 ---->

@@

Simulated standardized RRi time-dependent fluctuations based on prior parameter distributions are shown in [Figure 4](#fig-prior-sim).

<!---- FIGURE 4 ---->
![](figures/figure-4.png)
[**Figure 4**]{#fig-prior-sim}. (A) Simulated standardized RRi time-dependent fluctuations based on prior parameter distributions, illustrating predicted RRi responses to exercise. Shaded areas represent 95%, 80%, and 60% quantile CI, offering insight into expected physiological variability across parameters. (B) Prior distributions and 95% CI were used to generate predictions based on physiological constraints and graphical visualization of standardized RRi data.
<!---- FIGURE 4 ---->

@@

<!---- TABLE 2 ---->
| Parameter | Estimate^1^ | SE^1^ | Lower^2^ | Upper^2^ |
|-----|---:|---:|---:|---:|
| $\alpha$  |  861.78 | 5.73 |  850.57 |  872.85 |
| $\beta$   | -345.49 | 7.41 | -359.81 | -330.97 |
| $c$       |    0.84 | 0.01 |    0.82 |    0.86 |
| $\lambda$ |   -3.05 | 0.06 |   -3.16 |   -2.94 |
| $\phi$    |   -2.60 | 0.06 |   -2.71 |   -2.48 |
| $\tau$    |    6.71 | 0.05 |    6.61 |    6.81 |
| $\delta$  |    3.24 | 0.10 |    3.05 |    3.44 |
| $\sigma$  |   27.57 | 0.57 |   26.45 |   28.70 |

[**Table 2**]{#tbl-parameters}. Population-parameter values estimated from group-level analysis. ^1^ Estimates and SE are computed as the posterior distribution's median and median absolute deviation, respectively; ^2^ Lower and Upper bounds from the quantile-based CI~95%~ of the posterior distribution.
<!---- TABLE 2 ---->

@@
